---
title: "RCAS Generation and Concentration-Response Profiling: PPAR isoforms"
author: "Jesse Rogers"
date: "June 5, 2024"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
---

Extraction and analyis of Reference Class Associated Signatures (RCAS)
using U-2 OS and HepaRG chemical screening data. Specific isoforms of PPAR
(alpha, gamma, delta) are separated manually to investigate whether individual
isoforms may generate robust RCAS versus as a single class.

# Setup Environment

```{r env-setup, message=FALSE}
# import packages for visualization
# (note that other required packages will be imported by R scripts)
library(ggplot2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

# set directories (`dir_home` should be set to the top-level directory)
dir_home <- "~/NAM-integration/NAM-integration-pilot-packaged/"
dir_scripts <- paste0(dir_home, "scripts/")
setwd(dir_home)

# import scripts for RCAS generation and profiling
source(paste0(dir_scripts, "pipeline_refchem_assignment.R"))
source(paste0(dir_scripts, "pipeline_RCAS_generation.R"))
source(paste0(dir_scripts, "pipeline_RCAS_profiling.R"))

path_rcas <- c(
  paste0(dir_home, "data/examples/rcas_heparg_gene_bnd.RData"),
  paste0(dir_home, "data/examples/rcas_u2os_gene_bnd.RData")
)

# import rcas objects
composite_heparg <- selectRCASGenes(path_rcas[1])
composite_u2os <- selectRCASGenes(path_rcas[2])

th <- theme(text = element_text(family = "Helvetica"))
```

# Assign PPAR Agonists to Individual Reference Classes

The PPAR agonists found in existing `rcas` objects are divided using the 
RefChemDB S12 table. Specifically, the table is filtered for support>=5, and 
inclusion of PPAR as a target. Chemicals with multiple PPAR entries are assigned
to the isoform with greating support if there is an intended primary MoA, 
i.e. are drugs for a particular isoform.

```{r assign-refchems}
# get union of refchems from rcas objects
ref <- rbind(
  select(composite_heparg$httr.wide, dtxsid, ref_class),
  select(composite_u2os$httr.wide, dtxsid, ref_class)
)
ref <- distinct(ref, dtxsid, .keep_all = TRUE)

# import refchemdb and filter for support/multiple modes
refchemdb <- loadRefChemDB(
  homedir = dir_home,
  filename = "NIHMS1537541-supplement-Supplement1.xlsx",
  sheet = "S12 Data",
  fileurl = "https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6784312/bin/NIHMS1537541-supplement-Supplement1.xlsx"
)
refchem_filt <- filterSupport(refchemdb, support = 5)

# filter for PPAR agonists in ref + annotate with highest support label
refchem_ppar <- filter(
  refchem_filt,
  dsstox_substance_id %in% ref$dtxsid[ref$ref_class == "PPAR_Agonist"] &
  mode == "Positive"
)
# plot support levels per target
plot_refchem <- ggplot(
  refchem_ppar,
  aes(support, stringr::str_trunc(name, 30), fill = target_mode, label = support)
) +
  geom_col() +
  geom_text(
    position = position_stack(vjust = 0.5),
    family = "Helvetica"
  ) +
  labs(x = "RefChemDB Support", y = "Reference Chemical", fill = "Target") +
  theme_bw() +
  th
plot_refchem

# assign dtxsids based on relative majority
## DTXSID6040745 (L-165041) assigned to PPARD
## DTXSID4020290 (Pirinixic acid) assigned to PPARA
## DTXSID3029869 (Bezafibrate) assigned to PPARA
refchem_ppar <- refchem_ppar %>%
  group_by(dsstox_substance_id) %>%
  mutate(target_count = n()) %>%
  ungroup() %>%
  mutate(
    target_ppar = case_when(
      target_count == 1 ~ target_mode,
      dsstox_substance_id == "DTXSID6040745" ~ "PPARD_Positive",
      dsstox_substance_id == "DTXSID4020290" ~ "PPARA_Positive",
      dsstox_substance_id == "DTXSID3029869" ~ "PPARA_Positive"
    ),
    target_ppar = gsub("_Positive", "_Agonist", target_ppar)
  )
table(distinct(refchem_ppar, dsstox_substance_id, target_ppar)$target_ppar)

# assign new labels to ref
ref <- mutate(
  ref, ref_class = case_when(
    dtxsid %in% refchem_ppar$dsstox_substance_id ~ refchem_ppar$target_ppar[
      match(dtxsid, refchem_ppar$dsstox_substance_id)
    ],
    TRUE ~ ref_class
  )
)
# drop targets with <3 chemicals (i.e. PPARD/G)
ref <- filter(ref, !ref_class %in% c("PPARD_Agonist", "PPARG_Agonist"))
```

# Run RCAS with Isoform-Specific Labels

```{r run-rcas, eval=FALSE}
# specify paths for httr gene-level potency estimates
path_httr <- c(
  "/ccte/projects1/HTTr/screen_gene_cr/GENE_CR_heparg2d_toxcast_pfas_pe1_normal_0.05_conthits.RData",
  "/ccte/projects1/HTTr/screen_gene_cr/GENE_CR_u2os_toxcast_pfas_pe1_normal_0.05_conthits.RData"
)
# step through RCAS generation
httr <- loadData(path_httr[1], bmd.low.bnd = 0.1, bmd.up.bnd = 10)
httr.filtered <- filterData(httr, ref, col.type = "gene")
httr.wide <- widenData(httr.filtered, ref, col.type = "gene", metric.fill = 2.5)
aov.return <- calcANOVA(httr.wide)
posthoc.return <- calcPosthoc(httr.wide, aov.return$aov.p.adjust)
bmd <- filter(
  httr, gene %in% posthoc.return$posthoc.vars & dtxsid %in% httr.wide$dtxsid
) %>%
  mutate(ref_class = httr.wide$ref_class[match(dtxsid, httr.wide$dtxsid)])
rcas <- list(
  ref = ref,
  httr.wide = httr.wide,
  aov.estimate = aov.return,
  posthoc.estimate = posthoc.return,
  bmd = bmd
)
save(rcas, file = paste0(dir_home, "heparg_ppar/rcas_heparg_ppar.RData"))
```

```{r rcas-genes}
# conduct final gene selection + display summary table
composite <- selectRCASGenes(paste0(dir_home, "heparg_ppar/rcas_heparg_ppar.RData"))

composite$composite %>%
  count(class_1) %>%
  rename(class = class_1, size = n) %>%
  DT::datatable()

posthoc_ppar <- composite$posthoc.estimate$posthoc.sig %>%
  filter(class_1 == "PPARA_Agonist")

posthoc_ppar %>%
  rename(class = class_1) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3))) %>%
  arrange(mean_1) %>%
  DT::datatable()

# grab BMCs for genes identified towards PPAR in posthoc
httr_ppar <- composite$httr.wide %>%
  select(dtxsid, name, ref_class, all_of(unique(posthoc_ppar$variable))) %>%
  pivot_longer(
    all_of(unique(posthoc_ppar$variable)),
    names_to = "gene",
    values_to = "bmd_log"
  ) %>%
  mutate(ppar_class = case_when(
    ref_class == "PPARA_Agonist" ~ "PPARA_Agonist", TRUE ~ "Other Class"
  ))

# plot BMCs by ref_class
plot_bmc <- ggplot(
  httr_ppar, aes(bmd_log, gene, color = ppar_class)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    alpha = 0.8,
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  labs(x = "log10(BMC)", y = "Gene", color = "Reference\nClass") +
  theme_bw() +
  th
plot_bmc

plot_all <- cowplot::plot_grid(
  plot_refchem, plot_bmc,
  labels = c("A", "B"), ncol = 1, align = "v", axis = "lrtb"
)
ggsave(
  paste0(dir_home, "output/FigSX3.png"),
  width = 960, height = 960, units = "px", scale = 2
)
```

```{r reproducibility}
sessionInfo()
```