---
title: "RCAS Generation and Concentration-Response Profiling"
author: "Jesse Rogers"
date: "December 6, 2023"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
---

Extraction and analyis of Reference Class Associated Signatures (RCAS)
using U-2 OS and HepaRG chemical screening data. RCAS represent
unique gene sets more potent for individual classes of chemicals versus
those annotated for other modes-of-action.

# Setup Environment

```{r env-setup, message=FALSE}
# import packages for visualization
# (note that other required packages will be imported by R scripts)
library(ggplot2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)

# set directories (`dir_home` should be set to the top-level directory)
dir_home <- "~/NAM-integration/NAM-integration-pilot-packaged/"
dir_scripts <- paste0(dir_home, "scripts/")
setwd(dir_home)

# import scripts for RCAS generation and profiling
source(paste0(dir_scripts, "pipeline_refchem_assignment.R"))
source(paste0(dir_scripts, "pipeline_RCAS_generation.R"))
source(paste0(dir_scripts, "pipeline_RCAS_profiling.R"))
```

# Assign Reference Chemicals

Reference classes are developed using
[RefChemDB](https://doi.org/10.1186/s13040-022-00292-z) by
assigning chemicals to clusters of target-mode combinations. Chemicals are
assigned using a rule-based procedure:

-   Hierarchically cluster target_mode entries based on Jaccard distance 
    -   Determine clusters using a dendrogram height cutoff of 0.8 
-   For each chemical, count the number of target_mode entries that belong to 
each cluster 
-   Assign chemical to cluster with highest count 
    -   Settle ties using total support for each cluster (sum(support) for all 
target_mode entries in a cluster) 
    -   Remove remaining ties (equal target representation and support) 
-   Remove clusters with <3 chemicals assigned
-   Cluster labels assigned based on most frequent target_mode entries within
each cluster

```{r assign-refchems}
# assign reference chemicals from RefChemDB and export to file
path_refchems <- "data/examples/refchemdb_clusters.RData"
refchem_assign <- assignRefChems(path_refchems)

# summarize chemicals represented by clusters
refchem_assign %>%
    count(cluster, cluster_target) %>%
    arrange(desc(n)) %>%
    DT::datatable()
```

# Generate RCAS

## Conduct univariate potency analysis

Gene sets are generated for reference classes using High-Throughput
Transcriptomics (HTTr) screening data of 1,201 chemicals in U-2 OS and 
HepaRG cell lines. Gene-level potency estimates were previously generated from
DESeq2-moderated fold changes using the httrpathway package
[available on USEPA Github](https://github.com/USEPA/CompTox-httrpathway), and 
are used as the basis for determining gene sets.

`analyzeHTTrANOVA()` can take some time to run if many significant
genes are found (compiling HepaRG-based RCAS below takes ~10 minutes). Here, 
RCAS objects are pre-generated and saved in the `data/examples` directory.

```{r generate-rcas, eval=FALSE}
# specify paths for httr gene-level potency estimates
path_httr <- c(
  "/ccte/projects1/HTTr/screen_gene_cr/GENE_CR_heparg2d_toxcast_pfas_pe1_normal_0.05_conthits.RData",
  "/ccte/projects1/HTTr/screen_gene_cr/GENE_CR_u2os_toxcast_pfas_pe1_normal_0.05_conthits.RData"
)
path_rcas <- c(
    paste0(dir_home, "data/examples/rcas_heparg_gene.RData"),
    paste0(dir_home, "data/examples/rcas_u2os_gene.RData")
)

# generate RCAS for each dataset
rcas_heparg <- analyzeHTTrANOVA(
    filepath.data = path_httr[1],
    filepath.ref = path_refchems,
    col.httr.type = "gene",
    col.ref.class = "cluster_target",
    metric.fill = 2.5
)
save(rcas_heparg, file = path_rcas[1])

rcas_u2os <- analyzeHTTrANOVA(
    filepath.data = path_httr[2],
    filepath.ref = path_refchems,
    col.httr.type = "gene",
    col.ref.class = "cluster_target",
    metric.fill = 2.5
)
save(rcas_u2os, file = path_rcas[2])

# display results for significant genes: U-2 OS
rcas_u2os$posthoc.estimate$posthoc.sig %>%
    mutate(across(where(is.numeric), ~ signif(.x, 3))) %>%
    DT::datatable()
```

## Select final RCAS genes

Following ANOVA-based potency comparisons per gene, final gene sets are
constructed for each reference class by selecting genes whose BMD values are
significantly lower for a single class versus at least 70% of other classes.
RCAS are retained for each target if at least 10 genes meet this criterion.

Note that this selection step is performed within `profileRCAS()` and is done
here for visualizing RCAS sizes/potencies.

```{r compile-rcas}
# perform RCAS selection for each object
composite_heparg <- selectRCASGenes(path_rcas[1])
composite_u2os <- selectRCASGenes(path_rcas[2])

# combine rcas for each cell type
composite_all <- rbind(
    mutate(
        composite_heparg$composite,
        source = "heparg",
        n_classes = length(unique(composite_heparg$httr.wide$ref_class))
    ),
    mutate(
        composite_u2os$composite,
        source = "u2os",
        n_classes = length(unique(composite_u2os$httr.wide$ref_class))
    )
)

# compare sizes of retained RCAS
composite_all %>%
    count(source, class_1) %>%
    rename(class = class_1, size = n) %>%
    DT::datatable()
```

# Reproducibility

```{r reproducibility}
sessionInfo()
```